Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond

### 2024-04-08 main.tex
:pencil: Update channel shaping literature review

\begin{document}
\title{Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	% \thanks{
	% 	The authors are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{yang.zhao18, b.clerckx\}@imperial.ac.uk).
	% }
}
\maketitle

\begin{abstract}
	This paper investigates how a passive \gls{ris} can manipulate the \gls{mimo} point-to-point channel in terms of singular values and power gain.
	We depart from the widely-adopted diagonal phase shift model to a \gls{bd} architecture, which enables \emph{cooperative} wave scattering thanks to in-group connections between elements.
	An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise \emph{geodesic} \gls{rcg} in the Stiefel manifold, which performs multiplicative (instead of add-then-project) updates along the exponential map for faster convergence.
	% For the singular value shaping problem, we numerically characterize their Pareto frontiers by weighted sum optimization via geodesic \gls{rcg}, which generalizes most relevant metrics (e.g., degree of freedom and condition number).
	% Many analytical singular value bounds are also derived for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	For the singular value shaping problem, we characterize their Pareto frontiers by geodesic \gls{rcg} and derive analytical bounds for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	% For the singular value problem, we consider a numerical optimization approach that characterizes the Pareto frontiers as well as an analytical bounding approach in special (i.e., rank-deficient and fully-connected) scenarios.
	For the power gain maximization problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% Those channel shaping designs have a two-fold advantage.
	Those channel shaping designs have two major implications.
	First, they demystify the performance gain from off-diagonal entries in terms of \emph{subchannel rearrangement} and \emph{subspace alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in subchannel rearrangement and subspace alignment.
	Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} approach and a two-stage shaping-inspired approach.
	Simulation results indicate that the rate deficit from the latter diminishes as the \gls{bd}-\gls{ris} evolves towards unitary, while its shaping advantages over diagonal \gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
	This is empowered by the recent advances in \gls{ris}, a programmable metasurface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
	A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
	It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than relays, and greater scalability than \gls{mimo}.
	The most popular \gls{ris} application is \emph{joint beamforming} design with transceivers, which has attracted significant attention in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
	Since coherent scattering can be decomposed as equal-gain combining and transmission, \gls{ris} elements usually offer a squared asymptotic performance (e.g., second-order array gain and fourth-order power scaling law \cite{Zhao2022}) than transceiver antennas.
	% However, a large-dimensional \gls{ris} is generally required to compensate the power loss from double fading.
	% a second-order array gain and a fourth-order power scaling law has been reported for \gls{ris} elements \cite{Zhao2022} --- both achieves squared improvements than transceiver antennas.
	% since the coherent scattering can be viewed as equal-gain combining and transmission.
	On the other hand, \gls{ris} can also be employed for \emph{backscatter modulation} by periodically switching its reflection pattern.
	This creates a free-ride message stream (similar to index modulation \cite{Basar2017}) that can be either integrated into the legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022}, or exploited by surrounding low-power devices for energy-efficient uplink communications \cite{Liang2020,Zhao2022a,Yang2024}.

	% which has been investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.

	% which could be incorporated into the transmitter and receiver for \emph{beamforming}, employed as a free-rider information source for \emph{modulation}, or simply placed in space as a standalone device for \emph{channel shaping}.
	% Using \gls{ris} for {passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	% On the other hand, \gls{ris} also enables {backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Despite the great potentials and fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape the wireless channel?}
	The answer depends heavily on the scattering model and hardware architecture.
	In diagonal \gls{ris}, each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	This ideally translates to a diagonal scattering matrix with unit-magnitude entries, which only controls the phase of the scattered waves but cannot redistribute their amplitude.
	% This allows the \gls{ris} to control the phase of the scattered wave, but not its amplitude.
	The concept was soon generalized to \gls{bd}-\gls{ris} \cite{Shen2020a} which physically groups adjacent elements using passive reconfigurable components.\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices respectively.}
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate losslessly within the circuit and depart partially from any element in the same group.
	It can thus control both the amplitude and phase of the scattered wave, generalizing the scattering matrix to block-diagonal with unitary blocks.
	Such a powerful model can be realized at reduced hardware complexity using tree- and forest-connected architectures inspired by graph theory \cite{Nerini2024}.
	\gls{bd}-\gls{ris} can also function in hybrid transmitting-and-reflecting mode \cite{Li2023b} and multi-sector mode \cite{Li2023c} for full-space coverage and multi-user support.
	Many practical design challenges have been addressed including channel estimation \cite{Li2024}, mutual coupling \cite{Li2023f}, and wideband modelling \cite{Li2024a}.
	Its superiority has been studied extensively in \gls{siso} and \gls{miso} systems in terms of single-user \gls{snr} maximization \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and multi-user \gls{wsr} maximization \cite{Fang2023,Zhou2023,Li2023c,Soleymani2024}.
	When it comes to \gls{mimo}, most existing works are limited to very specific scenarios.
	For example, the authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design under the assumption of unitary \gls{ris} without direct link.
	% \gls{mimo} channel power maximization problem has been considered in \cite{Nerini2023,Nerini2023b}
	Received power maximization problem is extended to \gls{mimo} for continuous-valued \cite{Nerini2023} and discrete-valued \cite{Nerini2023b} scattering matrices, but it only considered single-stream transmission with given precoder and combiner (which is equivalent to \gls{siso}).
	Transmitter-side \gls{bd}-\gls{ris} design over statistical \gls{csi} has been introduced to massive \gls{mimo} networks for improved spectral efficiency \cite{Mishra2024}, but was again constrained to unitary \gls{ris} without direct link.
	A practical frequency-dependent \gls{bd}-\gls{ris} model has been recently proposed for multi-band multi-cell \gls{mimo} networks to facilitate practical deployments \cite{Sena2024}.
	Although the results are insightful, it remains unclear whether and why the advantage of \gls{bd}-\gls{ris} over diagonal \gls{ris} depends on \gls{mimo} dimensions.
	% , and there lacks a deep understanding of

	% Therefore, \gls{bd}-\gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	% Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself, allowing one to decouple \gls{ris} and transceiver design.
	Channel shaping is different from passive beamforming because \gls{ris} is exploited as a stand-alone device to modify the inherent properties of the wireless channel.
	This allows one to decouple the \gls{ris}-transceiver design and explore the fundamental limits of wave manipulation.
	Relevant design criteria can be roughly classified into two categories:
	\begin{itemize}
		\item \emph{Singular value centric:} Focus on the singular values of the equivalent channel matrix, which are closely related to the channel capacity and the diversity-multiplexing trade-off. Relevant designs are more complicated since they involve the entire channel matrix.
		\item \emph{Power centric:} Focus on the second-order characteristics of the channel matrix, which are useful in \gls{miso} but less informative in \gls{mimo}. Relevant designs are more practical and straightforward.
	\end{itemize}
	The former has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023} in \gls{pc}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023} in \gls{ic}.
	The latter has been studied in terms of channel power \cite{Wu2019,Nerini2023,Santamaria2023} in \gls{pc} and leakage interference \cite{Santamaria2023a} in \gls{ic}.
	Although the results are insightful, the big picture is far from complete since there exists neither a systematic channel shaping study nor a universal \gls{ris} optimization framework.
	To the best of our knowledge, the most general channel shaping scenario (\gls{bd}-\gls{ris}-aided \gls{mimo}) has not been studied, and the great potential of \gls{bd}-\gls{ris} has not been fully exploited or understood.
	This chapter aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	% Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself, allowing one to decouple \gls{ris} and transceiver design.
	Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself.
	This allows one to decouple the \gls{ris}-transceiver design and explore the fundamental limits of channel manipulation.
	For example, diagonal \gls{ris} has been proved useful for improving channel power \cite{Ning2020}, degree of freedom \cite{Ozdogan2020,Li2023h}, condition number \cite{Zheng2022,Huang2023}, and effective rank \cite{ElMossallamy2021,Meng2023} in \gls{mimo}.
	In contrast, \gls{bd}-\gls{ris} can provide a higher channel power but existing results are limited to \gls{siso}\footnote{In terms of channel shaping, single-stream \gls{mimo} with given precoder and combiner \cite{Nerini2023} is equivalent to \gls{siso}.}. \cite{Nerini2023} and \gls{miso} \cite{Santamaria2023}.
	While these studies offer promising glimpses into the channel shaping potential, a comprehensive understanding of the capabilities and limitations is desired, and a universal design framework is missing.
	This paper aims to answer the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

	First, we quantify the capability of a \gls{bd}-\gls{ris} to reshape the \gls{mimo} point-to-point channel in terms of singular values.
	The \emph{Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	% This generalizes most singular value metrics and provides a powerful design framework.
	The resulting singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	We then discuss some analytical singular value bounds in rank-deficient and fully-connected scenarios, which help to demystify the gain from off-diagonal entries.
	This is the first paper to answer the channel shaping question and highlight the \gls{bd}-\gls{ris} gain from a Pareto perspective.

	Second, we propose a \gls{rcg} algorithm adapted from \cite{Abrudan2008,Abrudan2009} for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size.
	Specifically, block-wise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold, which are expressed compactly by the exponential map \cite{Edelman1998}.
	It features lower complexity and faster convergence than general manifold optimization \cite{Absil2009,Pan2022d}, and solves the Pareto singular value problem.
	This is the first paper to tailor an efficient optimization framework for asymmetric \gls{bd}-\gls{ris}.

	Third, we tackle \gls{bd}-\gls{ris} \gls{mimo} rate maximization with two solutions: a local-optimal approach through \gls{ao} and a low-complexity approach over channel shaping.
	The former updates active and passive beamformers by eigenmode transmission and \gls{rcg} algorithm, respectively.
	The latter suboptimally decouples both blocks, recasts the shaping problem as channel power maximization, and solves it iteratively in closed form.
	Interestingly, the gap in between vanishes as \gls{bd}-\gls{ris} evolves from diagonal (single-connected) to unitary (fully-connected).
	It suggests channel shaping offers a promising low-complexity solution for joint \gls{ris}-transceiver designs.

@@

	Fourth, extensive simulations reveal that the performance gain from \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions.
	In terms of channel power, fully-connected \gls{bd}-\gls{ris} boosts up to 62\% and 270\% over diagonal \gls{ris} in $1 \times 1$ and $4 \times 4$ \gls{mimo} under independent Rayleigh fading, respectively.
	The superiority stems from stronger \emph{subchannel rearrangement} and \emph{subspace alignment} capabilities empowered by in-group cooperation.
	It emphasizes the importance of using \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems.

@@

\begin{section}{\gls{bd}-\gls{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis, we assume a lossless asymmetric network\footnote{While symmetric \gls{bd}-\gls{ris} (involving capacitors and inductors) is often considered in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2024,Zhou2023,Li2023d,Bartoli2023}, asymmetric \gls{bd}-\gls{ris} can be built over reconfigurable asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}.} without mutual coupling between scatter elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is a unitary matrix that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	That is, $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{H}$ and $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$.
	Note that diagonal (single-connected) and unitary (fully-connected) \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the radiation pattern and circuit topology have impacts on the scattering matrix.

>>>>

\begin{document}
\title{Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{yang.zhao18, c.li21, b.clerckx\}@imperial.ac.uk).

		Massimo Franceschetti is with the Department of Electrical and Computer Engineering, University of California at San Diego, La Jolla CA 92093, USA (e-mail: massimo@ece.ucsd.edu).
	}
}
\maketitle

\begin{abstract}
	This paper investigates how a passive \gls{ris} can manipulate the \gls{mimo} point-to-point channel in terms of singular values and power gain.
	We depart from the widely-adopted diagonal phase shift model to a \gls{bd} architecture, which enables {cooperative} wave scattering thanks to in-group connections between elements.
	An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise {geodesic} \gls{rcg} in the Stiefel manifold, which performs multiplicative (instead of add-then-retract) updates along the exponential map for faster convergence.
	% For the singular value shaping problem, we numerically characterize their Pareto frontiers by weighted sum optimization via geodesic \gls{rcg}, which generalizes most relevant metrics (e.g., degree of freedom and condition number).
	% Many analytical singular value bounds are also derived for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	For the singular value problem, we characterize their Pareto frontiers by geodesic \gls{rcg} and derive analytical bounds for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	% For the singular value problem, we consider a numerical optimization approach that characterizes the Pareto frontiers as well as an analytical bounding approach in special (i.e., rank-deficient and fully-connected) scenarios.
	For the power gain problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% Those channel shaping designs have a two-fold advantage.
	Those channel shaping designs have two major implications.
	First, they demystify the performance gain from off-diagonal entries in terms of {subchannel rearrangement} and {subspace alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in subchannel rearrangement and subspace alignment.
	Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} approach and a two-stage shaping-inspired approach.
	Simulation results indicate that the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal towards unitary, while the relative gain of \gls{bd}-\gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

@@

\begin{section}{Introduction}
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
	This is empowered by the recent advances in \gls{ris}, a programmable metasurface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
	A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
	It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than relays, and greater scalability than \gls{mimo}.
	The most popular \gls{ris} application is \emph{joint beamforming} design with transceivers, which has attracted significant attention in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
	Since coherent scattering can be decomposed as equal-gain combining and transmission, \gls{ris} elements usually offer a squared asymptotic performance (e.g., second-order array gain and fourth-order power scaling law \cite{Zhao2022}) than transceiver antennas.
	% However, a large-dimensional \gls{ris} is generally required to compensate the power loss from double fading.
	On the other hand, \gls{ris} can also be employed for \emph{backscatter modulation} by periodically switching its reflection pattern.
	This creates a free-ride message stream (similar to index modulation \cite{Basar2017}) that can be either integrated into the legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022}, or exploited by surrounding low-power devices for energy-efficient uplink communications \cite{Liang2020,Zhao2022a,Yang2024}.

	Despite the great potentials and fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape a \gls{mimo} channel?}
	The answer depends heavily on the scattering model and hardware architecture.
	In diagonal \gls{ris}, each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	This ideally translates to a diagonal scattering matrix with unit-magnitude entries, which only controls the phase of the scattered waves but cannot redistribute their amplitude.
	The concept was soon generalized to \gls{bd}-\gls{ris} \cite{Shen2020a} which physically groups adjacent elements using passive reconfigurable components.\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices respectively.}
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate losslessly within the circuit and depart partially from any element in the same group.
	It can thus control both the amplitude and phase of the scattered wave without power loss, generalizing the scattering matrix to block-diagonal with unitary blocks.
	Such a powerful model can be realized at reduced hardware complexity using tree- and forest-connected architectures inspired by graph theory \cite{Nerini2024}.
	\gls{bd}-\gls{ris} can also function in hybrid transmitting-and-reflecting mode \cite{Li2023b} and multi-sector mode \cite{Li2023c} for full-space coverage and multi-user support.
	Many practical design challenges have been addressed including channel estimation \cite{Li2024}, mutual coupling \cite{Li2023f}, and wideband modelling \cite{Li2024a}.
	Its superiority has been studied extensively in \gls{siso} and \gls{miso} systems in terms of single-user \gls{snr} maximization \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and multi-user \gls{wsr} maximization \cite{Fang2023,Zhou2023,Li2023c,Soleymani2024}.
	When it comes to \gls{mimo}, most existing works have been limited to very specific scenarios.
	For example, the authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design under the assumption of unitary \gls{ris} without direct link.
	Received power maximization problem is extended to \gls{mimo} for continuous-valued \cite{Nerini2023} and discrete-valued \cite{Nerini2023b} scattering matrices, but they only considered single-stream transmission with given precoder and combiner (which is equivalent to \gls{siso}).
	Transmitter-side \gls{bd}-\gls{ris} design over statistical \gls{csi} has been introduced to massive \gls{mimo} networks for improved spectral efficiency \cite{Mishra2024}, but was again constrained to unitary \gls{ris} without direct link.
	A practical frequency-dependent \gls{bd}-\gls{ris} model has been recently proposed for multi-band multi-cell \gls{mimo} networks to facilitate practical deployments \cite{Sena2024}.
	Although the results are insightful, it remains unclear how the off-diagonal entries contribute to channel shaping and whether the relative gain of \gls{bd}-\gls{ris} scales with \gls{mimo} dimensions.
	% Therefore, \gls{bd}-\gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	Channel shaping is different from passive beamforming since the \gls{ris} is exploited as a stand-alone device to modify the inherent properties of the channel itself, instead of optimizing for a specific performance metric.
	This not only allows one to decouple the joint \gls{ris}-transceiver design, but also unveils the fundamental limits of wave manipulation and provides an all-in-one benchmark for various wireless applications.
	Relevant designs can be classified into two categories:
	\begin{itemize}
		\item \emph{Singular value centric:} Focus on the singular values of the equivalent channel matrix, which are sensitive to minor perturbations but precisely related to the capacity and diversity-multiplexing trade-off. It has been mainly studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}.
		\item \emph{Power centric:} Focus on the second-order characteristics of the equivalent channel matrix, which are useful in \gls{siso} and \gls{miso} but less informative in \gls{mimo}. It has been mainly studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and leakage interference \cite{Santamaria2023a}.
	\end{itemize}

	While the above works offer promising glimpses into the channel shaping potential of passive \gls{ris}, the big picture remains incomplete because they neither suggest the achievable singular value region nor address the most general \gls{bd}-\gls{ris}-aided \gls{mimo} scenario.
	On the other hand, the great potential of \gls{bd}-\gls{ris} has not been fully understood and an efficient design framework is still missing.
	This chapter aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

	First, we exploit the Riemannian geometry of the Stiefel manifold to design a novel geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg} algorithm for \gls{bd}-\gls{ris} optimization.
	This method modified from \cite{Abrudan2008,Abrudan2009} not only features lower complexity and faster convergence than general non-geodesic approach \cite{Absil2009,Pan2022d}, but also fits arbitrary group size and any smooth optimization problem.
	Specifically, block-wise multiplicative rotational updates are performed along the geodesics of the Stiefel manifold, which are compactly evaluated as the exponential map \cite{Edelman1998}.
	By exploiting the inherent structure of unitary matrices, this method avoids retractions from the Euclidean space and improves the computational efficiency and stability.
	This is the first work to tailor an efficient and universal optimization framework for \gls{bd}-\gls{ris}.

	Second, we demystify the gain from off-diagonal entries of \gls{bd}-\gls{ris} in terms of subchannel rearrangement and subspace alignment.

	% Specifically,  we show that BD-RIS enables cooperative groups to rearrange subchannels according to their strengths, maximizing the effective channel gain. Additionally, intra-group cooperation allows for finer control over subspace alignment compared to diagonal RIS, especially in higher-dimensional MIMO systems. This analysis highlights a key advantage of BD-RIS and its potential for significant performance improvement.

	Third, we quantify the capability of a \gls{bd}-\gls{ris} to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The \emph{Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The resulting achievable singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and fully-connected (a.k.a. unitary) \gls{ris}.
	This is the first work to comprehensively answer the channel shaping question and fully unveil the potential of \gls{bd}-\gls{ris} in multi-antenna systems.

	Fourth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} achievable rate maximization with a local-optimal approach via \gls{ao} and a low-complexity approach over channel shaping.
	The former updates active and passive beamforming by eigenmode transmission and \gls{rcg} algorithm, respectively.
	The latter suboptimally decouples both blocks, recasts the shaping problem as channel power maximization, and solves it iteratively in closed form.
	Interestingly, the gap in between vanishes as \gls{bd}-\gls{ris} evolves from diagonal to unitary.
	It suggests channel shaping helps to decouple the \gls{ris}-transceiver design and offers a promising low-complexity solution.
	This is the first work to discuss the implications of channel shaping on joint active-passive beamforming optimization.

	Fifth, extensive simulations reveal that the performance gain from \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions.
	In terms of channel power, fully-connected \gls{bd}-\gls{ris} boosts up to 62\% and 270\% over diagonal \gls{ris} in $1 \times 1$ and $4 \times 4$ \gls{mimo} under independent Rayleigh fading, respectively.
	The superiority stems from stronger \emph{subchannel rearrangement} and \emph{subspace alignment} capabilities empowered by in-group cooperation.
	It emphasizes the importance of using \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems.

	% First, we quantify the capability of a \gls{bd}-\gls{ris} to reshape the \gls{mimo} point-to-point channel in terms of singular values.
	% The \emph{Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	% % This generalizes most singular value metrics and provides a powerful design framework.
	% The resulting singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	% We then discuss some analytical singular value bounds in rank-deficient and fully-connected scenarios, which help to demystify the gain from off-diagonal entries.
	% This is the first paper to answer the channel shaping question and highlight the \gls{bd}-\gls{ris} gain from a Pareto perspective.

	% Second, we propose a \gls{rcg} algorithm adapted from \cite{Abrudan2008,Abrudan2009} for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size.
	% Specifically, block-wise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold, which are expressed compactly by the exponential map \cite{Edelman1998}.
	% It features lower complexity and faster convergence than general manifold optimization \cite{Absil2009,Pan2022d}, and solves the Pareto singular value problem.
	% This is the first paper to tailor an efficient optimization framework for asymmetric \gls{bd}-\gls{ris}.

	% Third, we tackle \gls{bd}-\gls{ris} \gls{mimo} rate maximization with two solutions: a local-optimal approach through \gls{ao} and a low-complexity approach over channel shaping.
	% The former updates active and passive beamformers by eigenmode transmission and \gls{rcg} algorithm, respectively.
	% The latter suboptimally decouples both blocks, recasts the shaping problem as channel power maximization, and solves it iteratively in closed form.
	% Interestingly, the gap in between vanishes as \gls{bd}-\gls{ris} evolves from diagonal (single-connected) to unitary (fully-connected).
	% It suggests channel shaping offers a promising low-complexity solution for joint \gls{ris}-transceiver designs.

@@

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis, we assume a lossless asymmetric network\footnote{While symmetric \gls{bd}-\gls{ris} (involving capacitors and inductors) is often considered in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2024,Zhou2023,Li2023d,Bartoli2023}, asymmetric \gls{bd}-\gls{ris} can be built over reconfigurable asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}.} without mutual coupling between scatter elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is a unitary matrix that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	That is, $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{H}$ and $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$.
	Note that diagonal (single-connected) and unitary (fully-connected) \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the radiation pattern and circuit topology have impacts on the scattering matrix.
